\documentclass[11pt]{article}
\usepackage[width=7.0in, height=10.0in]{geometry}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{upgreek}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{tikz}



\newcommand{\vct}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\vr}{\vct{r}}
\newcommand{\vrN}{\vr^N}
\newcommand{\vrn}{\vr^n}
\newcommand{\dvr}{\frac{ d \vr  }{(2\pi)^3}}
\newcommand{\vx}{\vct{x}}
\newcommand{\vxN}{\vx^N}
\newcommand{\vxn}{\vx^n}
\newcommand{\dvx}{\frac{ d \vx  }{(2\pi)^3}}
\newcommand{\vk}{\vct{k}}
\newcommand{\dvk}{\frac{ d \vk  }{(2\pi)^3}}
\newcommand{\vy}{\vct{y}}
\newcommand{\vz}{\vct{z}}
\newcommand{\vE}{\vct{E}}
\newcommand{\vD}{\vct{D}}
\newcommand{\vP}{\vct{P}}
\newcommand{\vM}{\vct{M}}
\newcommand{\vMbar}{\overline{\vct{M}}}
\newcommand{\vdel}{\vct{\updelta}}
\newcommand{\vmu}{\vct{\upmu}}


\begin{document}


\title{The Feynman Lectures on Physics}
\author{ \vspace{-10ex} }
\date{ \vspace{-10ex} }
\maketitle



Here we review the Feynman's Lectures on Physics,
Vol. II Chapters 10 and 11.
%
We will use the SI units throughout.
If the reader who prefers the Gaussian units
should simply replace $\epsilon_0$ by $1/(4 \pi)$,
and ignore $4\pi \epsilon_0$ wherever it occurs.


\section{11-4 Electric fields in cavities of a dielectric}



It should be noted that the hole is
a ``virtual'' (Lorentz) cavity,
which is the negative of the plug.



\subsection{Models}



First, we recall that the Maxwell equations governing electrostatics
are linear with respect to the fields,
such as the electric field $\vE$ and the charge density $\rho$.
%
This simple fact is very useful,
for it allows us to build up a complex field
by linearly combining simpler electric fields.

In this section,
we will study a few basic model of polarization.
%
These models will be used as building blocks
to understand the more complex models
for the dielectric constant.



\subsection{Uniformly polarized ball}



The effect of a uniformly polarized ball of radius $R$ is equivalent to
that of a uniformly and positively charged ball
located at $\vct c_+ = (0, 0, \delta/2)$ (with charge density $\rho$)
and that of a uniformly and negatively charged ball
located at $\vct c_- = (0, 0, -\delta/2)$,
with $\delta \rightarrow 0$.

The electric field caused by the positive ball
can be computed using the Gauss theorem.
%
For $r_+ < R$, we have
\[
  E 4 \pi r_+^2
=
  \frac { 4 \pi r_+^3 } { 3 } \frac{ \rho } { \epsilon_0 },
\]
where
\[
  r_\pm = |\vr - \vct c_\pm| = r \mp \frac{ \delta \cos\theta} { 2 },
\]
with $\theta$ being the angle between $\vr$ and $\vz$,
i.e., $\hat\vr \cdot \hat\vz = \cos\theta$.
So
\[
  \vE_+
=
  \frac{ \rho } { 3 \, \epsilon_0 } ( \vr - \vct c_+ ).
\]
By symmetry the field caused by the negatively charged ball
is given by
\[
  \vE_- = -\frac{ \rho } { 3 \, \epsilon_0 } ( \vr - \vct c_- ).
\]
The superposition of the two yields
\begin{align}
  \vE_{\vP} = \vE_+ + \vE_-
= - \frac{ \rho \, \vdel } { 3 \epsilon_0 }
= - \frac{ \vP } { 3 \epsilon_0 }
= - \frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM } { R^3 },
  \label{eq:EP_in}
\end{align}
where $\vdel = \vz \, \delta$,
the polarization vector $\vP = \rho \vdel$,
%
and
\[
  \vM =
  \frac{ 4 \pi R^3 \, \rho \, \vdel } { 3 }
  =
  \frac{ 4 \pi R^3 } { 3 } \, \vP.
\]
is the total dipole moment of the ball.

Outside of the ball, we can also use the Gauss theorem.
\[
  E 4 \pi r_+^2
=
  \frac { 4 \pi R^3 } { 3 } \frac{ \rho } { \epsilon_0 },
\]
and
\[
  \vE_+
=
  \frac{ \rho \, R^3} { 3 \, \epsilon_0 \, r_+^3} ( \vr - \vct c_+ ).
\]
Similarly,
\[
  \vE_-
=
  -\frac{ \rho \, R^3} { 3 \, \epsilon_0 \, r_-^3} ( \vr - \vct c_- ).
\]
The sum of the above two is the electric field outside the ball.
For a small $\delta$,
we get
\begin{align}
  \vE_{\vP}
&=
  -\frac{ \rho \, R^3 \, (\vct c_+ - \vct c_-) } { 3 \, \epsilon_0 \, r^3 }
  +\frac{ \rho \, R^3 \vr } { 3 \, \epsilon_0 }
   \left(
      \frac{ 1 } { r_+^3 }
      -
      \frac{ 1 } { r_-^3 }
   \right)
   \notag \\
&=
  -\frac{ \rho \, R^3 \vdel } { 3 \, \epsilon_0 \, r^3 }
  +\frac{ \rho \, R^3 \vr } { \epsilon_0 }
      \frac{ \delta \cos\theta } { r^4 }
   \notag \\
&=
  -\frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM } {r^3}
  +
  \frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ 3 \, (\vM \cdot \hat\vr) \, \hat\vr } { r^3 }
   \notag \\
&=
  \frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM \cdot (3 \, \hat\vr \, \hat\vr - 1) } { r^3 }
=
  -\nabla \left(
  \frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM \cdot \vr } {r^3}
  \right).
  \label{eq:EP_out}
\end{align}


Physically,
%
A uniformly polarized ball produces,
within the ball,
a uniform field
that is $-1/\epsilon_0$ times the polarization vector,
by Eq. \eqref{eq:EP_in};
and
Eq. \eqref{eq:EP_out} says that
the electric field outside the ball
is equivalent to that of a dipole of moment $\vM$.


In summary,
we have
\begin{equation}
  \vE_{\vP}
=
  \begin{dcases}
  -\frac{ \vP } { 3 \, \epsilon_0 }
  =
  -\frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM } {R^3}
  & \mbox{for $r \le R$},
    \\
  -\nabla \left(
    \frac{ \vP \cdot \hat \vr \, R^3 } { 3 \, \epsilon_0 \, r^2 }
    \right)
  =
  -\nabla \left(
  \frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM \cdot \vr } {r^3}
  \right)
  & \mbox{for $r > R$}.
  \end{dcases}
  \label{eq:uball}
\end{equation}
%
Or in terms of the electric potential,
$\vE_{\vP} = -\nabla \phi_{\vP}$,
%
\begin{equation}
  \phi_{\vP}
=
  \begin{dcases}
  \frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM \cdot \vr } {R^3}
  & \mbox{for $r \le R$},
    \\
  \frac{ 1 } { 4 \pi \epsilon_0 }
  \frac{ \vM \cdot \vr } {r^3}
  & \mbox{for $r > R$}.
  \end{dcases}
  \label{eq:phi_uball}
\end{equation}



\subsection{Dielectric ball in uniform field}



First, let us consider a solid ball of radius $R$ with
the dielectric constant $\epsilon$.
The ball is placed in vacuum
under a constant electrostatic field
$\vE_0 = E_0 \vz$
along the $z$ axis.
We wish to know the electric field within the ball.

Solution.
Let us assume that the ball is uniformly polarized,
with polarization vector $\vP$.
Then the electric field caused by the polarization charges
is given by $-\vP/\epsilon_0$,
according to Eq. \eqref{eq:EP_in}.
The total electric field (for $r < R$) is given by
\[
  \vE = \vE_0 - \frac{ \vP } { 3 \, \epsilon_0 }.
\]
Further, we have
\[
  \frac{ \vP } { \epsilon_0 }
= (\epsilon - 1) \vE.
\]
From the two above equations, we get
\begin{align}
  \vE &= \frac{ 3 } { \epsilon + 2 } \vE_0,
  \label{eq:Eplug}
  \\
  \frac{\vP}{\epsilon_0}
  &= \frac{ 1 } { 4 \pi \epsilon_0 } \frac{ 3 \vM } { R^3 }
  = \frac{ 3 ( \epsilon - 1 ) } { \epsilon + 2 } \vE_0
  \label{eq:Pplug}
\end{align}



\subsection{Dielectric ball in uniform field, generalized}



Now consider a generalization.
%
The solid ball of dielectric constant $\epsilon$
is now immersed in a medium of dielectric constant $\epsilon'$.
%
A uniform electric field $\vE'$ is applied in the medium.

Solution.
The ball is still uniformly polarized,
but so is the medium.
Thus we shall consider instead
the excess polarization vector $\Delta\vP$
of the ball with respect to the medium.
%
Then the electric field caused by $\Delta\vP$
is given by $-\Delta\vP/\epsilon_0$,
according to Eq. \eqref{eq:EP_in}.
The total electric field inside the ball ($r < R$) is given by
\[
  \vE = \vE' - \frac{ \Delta \vP } { 3 \, \epsilon_0 \, \epsilon' }.
\]
Further, we have within the ball
\[
  \frac{ \Delta \vP } { \epsilon_0 }
  = \left( \epsilon - \epsilon' \right) \, \vE.
\]
Eliminating $\Delta \vP$ from the two above equations yields
\begin{equation}
  \vE = \frac{ 3 \, \epsilon' } {2 \, \epsilon' + \epsilon} \vE'.
  \label{eq:Eplug_general}
\end{equation}

Particularly for a vacuum cavity with $\epsilon = 1$,
we have
\begin{equation}
  \vE = \frac{ 3 \, \epsilon' } {2 \, \epsilon' + 1} \vE'.
  \label{eq:Ehole}
\end{equation}



\subsection{Alternative derivations of the cavity field}



\subsubsection{First alternative derivation}



Now let us use \eqref{eq:Eplug} to derive the cavity field
\[
  \vE_{\mathrm{hole}}
=
  \frac { 3 } { \epsilon + 2 }
  \vE.
\]


\begin{enumerate}
\item
  Consider a plug of dielectric constant $\epsilon$
  placed in vacuum under the electric field $\vE_0$.
  According to Eq. \eqref{eq:Eplug},
  the plug attains a uniform polarization
  and the field strength is
  \[
    \vE_1 = \frac { 3 } { \epsilon + 2 } \vE_0.
  \]

\item
  We know that in the dielectric,
  the plug achieves the same polarization
  under the uniform field $\vE$.
  Thus
  \[
    \vE = \vE_1 = \frac { 3 } { \epsilon + 2 } \vE_0.
  \]

\item
  The field $\vE = \vE_1$ is the sum of the external field $\vE_0$
  and the field caused by the polarization of the plug $\vE_{\vP}$,
  So we have
  \[
    \vE_{\vP}
    =
    \vE - \vE_0
    =
    -\frac{ \epsilon - 1 } { \epsilon + 2 } \vE_0.
    =
    -\frac{ \epsilon - 1 } { 3 } \vE.
  \]

\item
  This field $\vE_{\vP}$ should be deducted for the hole or the virtual cavity.
  Thus, the total field within the cavity is given by
  \[
    \vE_\texttt{hole}
    =
    \vE - \vE_{\vP}
    =
    \frac{ \epsilon + 2 } { 3 } \vE
    =
    \vE_0.
  \]

\end{enumerate}



\subsubsection{Second alternative derivation}



The simple conclusion $\vE_{hole} = \vE_0$
calls for a simpler derivation.
Below, we give another derivation based on
the superposition principle.


\begin{equation}
\newcommand{\sz}{4mm}
\newcommand{\szR}{4.0*\sz}
\newcommand{\szr}{1.5*\sz}
\newcommand{\bs}{0mm}
{
  \begin{tikzpicture}[baseline=\bs]
    \fill[yellow!40!white] (0, 0) circle (\szR);
    \fill[yellow!5!white] (0, 0) circle (\szr);
  \end{tikzpicture}
}
\,=\,
{
  \begin{tikzpicture}[baseline=\bs]
    \fill[yellow!40!white] (0, 0) circle (\szR);
  \end{tikzpicture}
}
\,-\,
{
  \begin{tikzpicture}[baseline=\bs]
    \fill[yellow!40!white] (0, 0) circle (\szr);
  \end{tikzpicture}
}\,.
\end{equation}


\begin{enumerate}

\item
The virtual cavity is constructed
by subtracting a small plug from the large plug.
The large plug is also spherical,
it represents the entire sample.

\item
Under an external (vacuum) field $\vE_0$,
both plugs attain the same electric field within.
The polarization vectors,
and the electric field $\vE_{\vP}$
caused by polarization
in the two plugs are also the same.

\item
Therefore, by subtracting the two plugs,
the electric field $\vE_{\vP}$
caused by polarization
is zero.

\item
Thus, the total electric field
within the hole is equal to the external field
$\vE_0$.

\end{enumerate}



\bibliography{../../liquid}
\bibliographystyle{alpha}
\end{document}
